One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2015-01-09 paper/discussion.tex
second rough draft of revised paper

{\bf
  \subsection{Motion of the central star} 
  In our models we have purposefully neglected the indirect potential
  associated with a non-inertial reference frame to avoid
  complications arising from the motion of the central star.  
  It has been established that such motion, in response to an $m=1$
  disturbance in the disc, can render the disturbance
  unstable in a barotropic disc (so the background torque is
  ineffective).

  However, for the disc masses in our models, $M_d\lesssim 
  0.1 M_*$, this effect is not expected to be significant
  \citep{adams89,shu90,michael10}. Indeed, simulations including the
  indirect potential, carried out in the early stages of this project,
  produced similar results.

@@

  Another possibility is a gap opened by an embedded planet. In that case $Q$
  rises rapidly towards the gap centre since it is a region of low
  surface density. This can satisfy Eq. \ref{qb_cond}. Then the inner
  edge of our bump function mimics the outer gap edge. The outer gap
  edge is then a potential site for the growth of a low-frequency
  one-armed spiral through the background torque. However, the
  locally isothermal requirement limits this process to the
  outer disc.   
}

>>>>

{\bf
  \subsection{Motion of the central star} 
  In our models we have purposefully neglected the indirect potential
  associated with a non-inertial reference frame to avoid
  complications arising from the motion of the central star.  
  Although it has been established that such motion can destabilise an
  $m=1$ disturbance in the disc \citep{adams89,shu90,michael10}, 
  the disc masses in our models ($M_d\lesssim 
  0.1 M_*$) are not expected to be sufficiently massive for this effect to be
  significant. Indeed, simulations including the
  indirect potential, carried out in the early stages of this project,
  produced similar results.

@@

  Another possibility is a gap opened by an embedded planet. In that case $Q$
  rises rapidly towards the gap centre since it is a region of low
  surface density. This can satisfy Eq. \ref{qb_cond}. Then the inner
  edge of our bump function mimics the outer gap edge. The outer gap
  edge is then a potential site for the growth of a low-frequency
  one-armed spiral through the background torque. However, the
  locally isothermal requirement would limit this process to the
  outer disc irradiated by the central star, or that the temperature
  profile about the gap edge is set by the planet luminosity.  
}

### 2015-01-09 paper/model.tex
second rough draft of revised paper

\section{Linear density waves}\label{wkb}
% In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
% stability of disc equilibria described 
% above. However, it is useful to  interpret simulation results
% using linear density theory. We list here the main definitions and
% results from 2D linear theory, taken from \cite{lin11b} and
% \cite{shu91}.  
{\bf We highlight a key feature of locally isothermal discs that
  enables angular momentum exchange between small disturbances and the
  background disc through an imposed radial temperature gradient. This
  conclusion results from the consideration of angular momentum
  conservation within the framework of linear perturbation theory. 
  For simplicity, we consider 2D discs, although generalisation to 3D is straight
  forward.  
}

In a linear analysis, one assumes a steady axisymmetric background state, % such as
% that defined in \S\ref{model},
which is then perturbed such that
\begin{align}  
  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
        m\phi\right)\right]}, 
\end{align}
and similarly for other variables, where $\sigma=\omega+\ii\gamma$ is
a complex frequency with $\omega$ being the real frequency, 
$\gamma$ the growth rate, and $m$ is an integer. Here, we take $m>0$ without
loss of generality.  
%Since the background is
%axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
 %   t\right)} = \Sigma_m$ as defined for diagnostics.

@@

{\bf 
  We demonstrate instability for low-frequency
  ($\omega\ll m\Omega$) perturbations by explicitly calculating its 
  angular momentum density 
  $\jlin$ and the background torque $T_\mathrm{BG}$, and show that
  they are unstable when a negative radial 
  temperature gradient is imposed.  
}

@@

Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
\omega/m$ of a neutral mode 
, Eq. \ref{dispersion} can be solved 
for $|k|$, 
\begin{align}\label{wavenumber}
%  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
 %     Q^2(1-\nu^2)}\right], 
|k| = k_c\left[1 \pm \sqrt{1 -
     Q^2(1-\nu^2)}\right], 
\end{align}
where 
 \begin{align}
   k_c \equiv \frac{\pi G \Sigma}{c_s^2}
 \end{align}
 is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. In
Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
spirals.

@@

This torque density is negative for trailing waves in discs with
a fixed temperature profile decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
respectively). Note that this conclusion does not rely on the
low-frequency approximation and holds for any $m$.

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadratic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for gives
\begin{align}\label{theoretical_rate0}
  2\gamma \sim
  -\frac{dc_s^2}{dR}
  \frac{k}{\Omega}\frac{|\xi_R|^2}{\left[(m-1)|\bm{\xi}|^2 + |\xi_R +
      \ii\xi_\phi|^2 \right]} 
  = -\frac{dc_s^2}{dR}
  \frac{k}{m\Omega},
%  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
%   \xi_\phi|^2}. 
\end{align}
{\bf where the second equality uses $\xi_\phi \simeq
2\ii \xi_R/m$ in the local approximation, as shown in  Appendix
\ref{horizontal_displacements}.} 
 % {\bf
% This holds for local modes of all frequencies and can be shown
% explicitly by evaluating Eq. \ref{ang_mom_def} with 
% \begin{align}
%   \xi_\phi = -\frac{2\ii\Omega}{\sbar}\xi_R, 
% \end{align}
% consistent with the local approximation as shown in Appendix
% \ref{horizontal_displacements}. 
% }
Then 
%To further simplify this expression, we need to relate $\xi_R$ and
%$\xi_\phi$. 
% \begin{align}
%   2\gamma \sim
%   -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
%     + \ii\xi_\phi|^2}.
% \end{align}
% Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
% in the local approximation, we have
% \begin{align}
%   2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
%     \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
%   \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
% \end{align}
% where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
% inequality. 
%where we used $dc_s^2/dR <0$ and $k>0$ to obtain the inequality. 
for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{m\Omega} \sim q h
  \left(\frac{kH}{m}\right)\Omega,\label{theoretical_rate}
\end{align}
where we used $c_s\sim H\Omega\sim hR\Omega$. 
{\bf It is important to note that, although our description of the local
  approximation accounts for disc self-gravity, mode growth due 
  to the fixed temperature gradient does not depend explicitly on
  whether or not the disc is self-gravitating, i.e. destabilisation is
  not due to self-gravity.}

>>>>

\section{Linear density waves}\label{wkb}
% In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
% stability of disc equilibria described 
% above. However, it is useful to  interpret simulation results
% using linear density theory. We list here the main definitions and
% results from 2D linear theory, taken from \cite{lin11b} and
% \cite{shu91}.  
{\bf We highlight a key feature of locally isothermal discs that
  enables angular momentum exchange between small disturbances and the
  background disc through an imposed radial temperature gradient. This
  conclusion results from the consideration of angular momentum
  conservation within the framework of linear perturbation theory. 
  For simplicity, in this section we consider 2D discs. 
}

In a linear analysis, one assumes a steady axisymmetric background state, % such as
% that defined in \S\ref{model},
which is then perturbed such that
\begin{align}  
  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
        m\phi\right)\right]}, 
\end{align}
and similarly for other variables, where $\sigma=\omega+\ii\gamma$ is
a complex frequency with $\omega$ being the real frequency, 
$\gamma$ the growth rate, and $m$ is an integer. We take $m>0$ without
loss of generality.  
%Since the background is
%axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
 %   t\right)} = \Sigma_m$ as defined for diagnostics.

@@

{\bf 
  We demonstrate instability for low-frequency modes 
  ($|\omega|\ll m\Omega$) by explicitly showing its 
  angular momentum density 
  $\jlin<0$, and the background torque $T_\mathrm{BG}<0$ for
  appropriate perturbations and radial temperature gradients. 
%, and show that
%  low-frequency modes can be unstable when a radial 
%  temperature gradient is imposed.  
}

@@

Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
\omega/m$ of a neutral mode 
, Eq. \ref{dispersion} can be solved 
for $|k|$, 
\begin{align}\label{wavenumber}
%  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
 %     Q^2(1-\nu^2)}\right], 
|k| = k_c\left[1 \pm \sqrt{1 -
     Q^2(1-\nu^2)}\right], 
\end{align}
where 
 \begin{align}
   k_c \equiv \frac{\pi G \Sigma}{c_s^2}
 \end{align}
 is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. In
Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
waves.

@@

This torque density is negative for trailing waves ($k>0$) in discs with
a fixed temperature profile decreasing outwards ($dc_s^2/dR<0$). Note
that this conclusion does not rely on the low-frequency approximation.

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadratic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for gives
\begin{align}\label{theoretical_rate0}
  2\gamma \sim
  -\frac{dc_s^2}{dR}
  \frac{k}{\Omega}\frac{|\xi_R|^2}{\left[(m-1)|\bm{\xi}|^2 + |\xi_R +
      \ii\xi_\phi|^2 \right]} 
  = -\frac{dc_s^2}{dR}
  \frac{k}{m\Omega},
%  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
%   \xi_\phi|^2}. 
\end{align}
{\bf where the second equality uses $\xi_\phi \simeq
2\ii \xi_R/m$ in the local approximation, as shown in  Appendix
\ref{horizontal_displacements}.} 
 % {\bf
% This holds for local modes of all frequencies and can be shown
% explicitly by evaluating Eq. \ref{ang_mom_def} with 
% \begin{align}
%   \xi_\phi = -\frac{2\ii\Omega}{\sbar}\xi_R, 
% \end{align}
% consistent with the local approximation as shown in Appendix
% \ref{horizontal_displacements}. 
% }
Then 
%To further simplify this expression, we need to relate $\xi_R$ and
%$\xi_\phi$. 
% \begin{align}
%   2\gamma \sim
%   -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
%     + \ii\xi_\phi|^2}.
% \end{align}
% Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
% in the local approximation, we have
% \begin{align}
%   2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
%     \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
%   \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
% \end{align}
% where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
% inequality. 
%where we used $dc_s^2/dR <0$ and $k>0$ to obtain the inequality. 
for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{m\Omega} \sim q h
  \left(\frac{kH}{m}\right)\Omega,\label{theoretical_rate}
\end{align}
where we used $c_s\sim H\Omega\sim hR\Omega$. 
{\bf Note the absence of the Toomre parameter $Q$ in
  Eq. \ref{theoretical_rate}. Although our description of the local 
  approximation accounts for disc self-gravity, mode growth due 
  to the fixed temperature gradient does not depend explicitly on
  whether or not the disc is self-gravitating, i.e. destabilisation is
  not due to self-gravity.}

### 2015-01-09 paper/paper.tex
second rough draft of revised paper

\maketitle
\begin{abstract} 
  {\bf We describe a new mechanism that leads to the destabilisation of
 non-axisymmetric waves in astrophysical discs with 
 an imposed radial temperature gradient. This might apply, for
 example, to the outer parts of protoplanetary discs. We use linear
 density wave theory 
 to show that non-axisymmetric perturbations generally do not conserve
 their angular momentum in the presence of a forced temperature
 gradient. This implies an exchange of angular momentum
 between linear perturbations and the background disc. In particular, 
 when the disturbance is a low-frequency
 trailing wave and the disc temperature decreases outwards, this 
 interaction is unstable and leads to the growth of the wave.  
 We demonstrate this phenomenon through numerical hydrodynamic
 simulations of locally isothermal discs in 2D using the FARGO
 code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
 radially structured discs with a self-gravitating region which remains
 stable in the absence of a temperature gradient. However,
 when a temperature gradient is imposed we observe exponential growth of an one-armed  
 spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
 outside the bulk of the spiral arm, resulting in a nearly-stationary
 one-armed spiral pattern.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by the imposed temperature gradient does 
 not explicitly require disc self-gravity, we suggest this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs. 
}
\end{abstract}

@@

{\bf We employ direct hydrodynamic simulations using three different
  grid-based codes to demonstrate how this `background torque' 
  can lead to the growth of one-armed spirals in radially structured, 
  self-gravitating discs. This is despite the fact that our disc
  models do not meet the requirements for the `SLING'
  mechanism. Although our numerical simulations consider
  self-gravitating discs, this `background torque' is generic for
  locally isothermal discs and its existence does not depend
  explicitly on disc self-gravity. Thus, the destabilisation effect we
  describe should also be applicable to non-self-gravitating discs.  
}

This paper is organised as follows. We describe the system of interest
and list the governing equations in 
\S\ref{model}. {\bf In \S\ref{wkb} we use linear theory  to show
  how a fixed temperature profile 
  can destabilise non-axisymmetric waves in discs.} \S\ref{methods} describes
the {\bf numerical setup and} hydrodynamic codes {\bf we use to 
  demonstrate the growth of one-armed spirals due to an imposed
  temperature gradient.} Our {\bf
  simulation} results are presented in  \S\ref{results2d} for two-
dimensional (2D) {\bf
  discs} and \S\ref{results3d} for three-dimensional (3D) {\bf discs}, and we further
discuss them in \S\ref{discussions}. We summarise in \S\ref{summary} 
with a discussion on speculations for future work.

@@

{\bf
\section*{Acknowledgments}
This material is based upon work supported by the National Science Foundation under Grant No. 1228509.
}
\bibliographystyle{mn2e}
\bibliography{ref}

>>>>

\maketitle
\begin{abstract} 
  {\bf We describe a new mechanism that leads to the destabilisation of
 non-axisymmetric waves in astrophysical discs with 
 an imposed radial temperature gradient. This might apply, for
 example, to the outer parts of protoplanetary discs. We use linear
 density wave theory 
 to show that non-axisymmetric perturbations generally do not conserve
 their angular momentum in the presence of a forced temperature
 gradient. This implies an exchange of angular momentum
 between linear perturbations and the background disc. In particular, 
 when the disturbance is a low-frequency
 trailing wave and the disc temperature decreases outwards, this 
 interaction is unstable and leads to the growth of the wave.  
 We demonstrate this phenomenon through numerical hydrodynamic
 simulations of locally isothermal discs in 2D using the FARGO
 code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
 radially structured discs with a self-gravitating region which remains
 stable in the absence of a temperature gradient. However,
 when a temperature gradient is imposed we observe exponential growth of an one-armed  
 spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
 outside the bulk of the spiral arm, resulting in a nearly-stationary
 one-armed spiral pattern.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by the imposed temperature gradient does 
 not explicitly require disc self-gravity, this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs, e.g in Be star discs. 
}
\end{abstract}

@@

{\bf We employ direct hydrodynamic simulations using three different
  grid-based codes to demonstrate how this `background torque' 
  can lead to the growth of one-armed spirals in radially structured, 
  self-gravitating discs. This is despite the fact that our disc
  models do not meet the requirements for the `SLING'
  mechanism. Although our numerical simulations consider
  self-gravitating discs, this `background torque' is generic for
  locally isothermal discs and its existence does not require
  disc self-gravity. Thus, the destabilisation effect we
  describe should also be applicable to non-self-gravitating discs.  
}

This paper is organised as follows. In \S\ref {model} we describe the system of interest
and list the governing equations. {\bf In \S\ref{wkb} we use linear theory  to show
  how a fixed temperature profile 
  can destabilise non-axisymmetric waves in discs.} \S\ref{methods} describes
the {\bf numerical setup and} hydrodynamic codes {\bf we use to 
  demonstrate the growth of one-armed spirals due to an imposed
  temperature gradient.} Our {\bf
  simulation} results are presented in  \S\ref{results2d} for two-dimensional (2D) {\bf
  discs} and in \S\ref{results3d} for three-dimensional (3D) {\bf discs}, and we further
discuss them in \S\ref{discussions}. We summarise in \S\ref{summary} 
with a discussion on speculations for future work.

@@

{\bf
\section*{Acknowledgments}
I thank K. Kratter and A. Youdin for valuable discussions, and the
anonymous referee for comments that significantly improved this paper. All
computations were performed on the El Gato cluster at the University
of Arizona. This material is based upon work supported by the National
Science Foundation under Grant No. 1228509. 
}
\bibliographystyle{mn2e}
\bibliography{ref}

### 2015-01-09 paper/summary.tex
second rough draft of revised paper

In the local approximation, we showed that this background torque is
negative for {\bf non-axisymmetric} trailing waves in discs with a fixed temperature or
sound-speed profile that decrease outwards. {\bf The background torque} enforces  
low-frequency {\bf non-axisymmetric} modes because they are associated
with negative angular momentum.

@@

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to show that the growth of     
one-armed spirals in our disc model is due to the imposed
temperature {\bf gradient: growth rates increased linearly}  
with the magnitude of the imposed temperature gradient, and one-armed
spirals did not develop in strictly isothermal simulations. {\bf This 
  one-armed spiral instability can be interpreted as an initially 
  neutral, tightly-wound $m=1$ mode being destabilised by the 
  background torque.} The spiral
is mostly confined between two $Q$-barriers surface density bump. 
We find the instability behaves similarly in 2D and 3D {\bf, but in 3D}
the spiral disturbance becomes more radially global away from    
the midplane.

@@

@@

{\bf The background torque in other contexts.}  
{\bf In our simulations}, self-gravity coupled with the disc 
structure permits a neutral one-armed mode, which is then
destabilised by the {\bf fixed} temperature gradient.  
Other origins of the neutral one-armed spiral should be
investigated. One possibility is in Be star discs \citep{rivinius13},
for which one-armed oscillations may explain long-timescale variations
in their emission lines \citep[see e.g.][and references 
therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
strictly isothermal disc models, so it would be interesting to explore
the effect of a temperature gradient on the stability of these
oscillations.

>>>>

In the local approximation, we showed that this background torque is
negative for {\bf non-axisymmetric} trailing waves in discs with a fixed temperature or
sound-speed profile that decrease outwards. {\bf A negative background torque} enforces  
low-frequency {\bf non-axisymmetric} modes because they are associated
with negative angular momentum.

@@

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to show that the growth of     
one-armed spirals in our disc model is due to the imposed
temperature {\bf gradient: growth rates increased linearly}  
with the magnitude of the imposed temperature gradient, and one-armed
spirals did not develop in strictly isothermal simulations. {\bf This 
  one-armed spiral instability can be interpreted as an initially 
  neutral, tightly-wound $m=1$ mode being destabilised by the 
  background torque.} The spiral
is mostly confined between two $Q$-barriers in the surface density bump. 
We find the instability behaves similarly in 2D and 3D{\bf, but in 3D}
the spiral disturbance becomes more radially global away from    
the midplane.

@@

%interaction with turbulence

@@

{\bf\emph{Other applications of the background torque.}}     
{\bf The background torque is a generic feature in  
  discs for which the temperature is set externally.  It may therefore
  be relevant in other astrophysical contexts. }
One possibility is in Be star discs \citep{rivinius13}, 
for which one-armed oscillations may explain long-timescale variations
in their emission lines \citep[see e.g.][and references 
therein]{okasaki97,papaloizou06c,ogilvie08}. {\bf These studies 
invoke other mechanisms to produce \emph{neutral} one-armed
oscillations (e.g. rotational deformation of the star), but consider 
strictly isothermal discs.} It would be interesting to explore
the effect of a radial temperature gradient on the stability of these
oscillations.
